---
title: "data3888_CM1_draft"
format: html
editor: visual
---

```{r}
library(arrow)
library(dplyr)

ds <- open_dataset("/Users/vaniakumar/Desktop/data3888/data3888_volatility_modelling/individual_book_train_parquet")
```

same rows for more than 1 col

using WAP to find problems

```{r}
colnames(ds)
```

```{r}
ds$schema
```

ds_filtered \<- ds %\>%

filter(

bid_price1 \> 0,

ask_price1 \> 0,

bid_price1 \< ask_price1

)

```{r}
library(arrow)
library(dplyr)

# Open the dataset
ds <- open_dataset("/Users/vaniakumar/Desktop/data3888/data3888_volatility_modelling/individual_book_train_parquet")

# Get all stock IDs
stock_ids <- ds %>% distinct(stock_id) %>% collect() %>% pull(stock_id)

# Loop over each stock and process separately
for (sid in stock_ids) {
  
  # 1️⃣ Filter, compute WAP and spreads (Arrow can handle this)
  stock_clean <- ds %>%
    filter(
      stock_id == sid,
      bid_price1 > 0,
      ask_price1 > 0,
      bid_price1 < ask_price1,
      bid_size1 + ask_size1 != 0
    ) %>%
    mutate(
      wap = (bid_price1 * ask_size1 + ask_price1 * bid_size1) / (bid_size1 + ask_size1),
      spread = ask_price1 - bid_price1,
      rel_spread = (ask_price1 - bid_price1) / ((bid_price1 + ask_price1)/2)
    ) %>%
    arrange(time_id, seconds_in_bucket) %>%
    collect()  # pull into R memory for this stock only

  # 2️⃣ Compute log returns after collecting
  stock_clean <- stock_clean %>%
    mutate(
      log_return = log(wap / lag(wap))
    )
}
```

```{r}
library(dplyr)
library(arrow)

ds <- open_dataset("/Users/vaniakumar/Desktop/data3888/data3888_volatility_modelling/individual_book_train_parquet")

# Get all stock IDs
stock_ids <- ds %>% distinct(stock_id) %>% collect() %>% pull(stock_id)

# Loop over each stock
for (sid in stock_ids) {
  
  cat("\n====================\nStock", sid, "\n====================\n")
  
  # 1 Filter, compute WAP and spreads
  stock_clean <- ds %>%
    filter(
      stock_id == sid,
      bid_price1 > 0,
      ask_price1 > 0,
      bid_price1 < ask_price1,
      bid_size1 + ask_size1 != 0
    ) %>%
    mutate(
      wap = (bid_price1 * ask_size1 + ask_price1 * bid_size1) / (bid_size1 + ask_size1),
      spread = ask_price1 - bid_price1,
      rel_spread = (ask_price1 - bid_price1) / ((bid_price1 + ask_price1)/2)
    ) %>%
    arrange(time_id, seconds_in_bucket) %>%
    collect()  # only pull this stock into memory

  # 2 Compute log returns
  stock_clean <- stock_clean %>%
    mutate(
      log_return = log(wap / lag(wap)),
      is_zero = log_return == 0
    )

  # --------- CASE 1: consecutive zero log returns ---------
  zero_runs <- rle(stock_clean$is_zero)
  zero_lengths <- zero_runs$lengths[zero_runs$values == TRUE]
  cat("Case 1: Consecutive zero log returns\n")
  print(summary(zero_lengths))  # see if there are long sequences
  if(max(zero_lengths, na.rm = TRUE) > 10){
    cat("  ⚠️ There are very long stretches of zero log returns!\n")
  } else {
    cat("  No unusual long stretches of zeros.\n")
  }
  
  # --------- CASE 2: huge spikes in log returns ---------
  spikes <- stock_clean %>% filter(abs(log_return) > 0.05)  # adjust threshold
  cat("Case 2: Huge log return spikes\n")
  print(summary(spikes$log_return))
  if(nrow(spikes) > 0){
    cat("  ⚠️ Some spikes detected.\n")
  } else {
    cat("  No spikes detected.\n")
  }

  # --------- CASE 3: abnormal relative spreads ---------
  wide_spreads <- stock_clean %>% filter(rel_spread > 0.1)
  cat("Case 3: Abnormal relative spreads (>10%)\n")
  print(summary(wide_spreads$rel_spread))
  if(nrow(wide_spreads) > 0){
    cat("  ⚠️ Some unusually wide spreads detected.\n")
  } else {
    cat("  No wide spreads detected.\n")
  }
  
  # --------- CASE 4: tiny WAP change but large spread ---------
  suspicious <- stock_clean %>% filter(abs(wap - lag(wap)) < 1e-5 & spread > 0.05)
  cat("Case 4: Tiny WAP change but large spread\n")
  print(head(suspicious, 10))  # show first few rows
  if(nrow(suspicious) > 0){
    cat("  ⚠️ Some suspicious rows detected.\n")
  } else {
    cat("No suspicious rows detected.\n")
  }
  
  cat("\n-------------------------------------\n")
}
```

```{r}
View(wide_spreads)
```
