---
title: "Untitled"
output: html_document
date: "2026-04-22"
editor_options: 
  markdown: 
    wrap: 72
---

# WARNING:

Ask about "what do you mean by "no longer need to manually type
user_features because ... auto-generate them all" ? As in the features
have already been optimized ? Cause the original plan was to do this
manually. Or at least give the option to allow user to manaully select
features.

log returns multiplied by 10k to avoid floating point underflows

Defined Var as mean of squared returns.

```{r, message=FALSE, warning=FALSE}
library(arrow)
library(dplyr)
library(purrr)
library(ranger)
library(rugarch)
library(zoo)
```

```{r}
#parquet
parquet_dir <- "individual_book_train_parquet"
ds <- open_dataset(parquet_dir, format = "parquet")
```

```{r}
#performance metric storage
pm_rmse <- function(actual_var, predicted_var) {
  return(sqrt(mean((actual_var - predicted_var)^2, na.rm = TRUE)))
}
pm_qlike <- function(actual_var, predicted_var) {
  predicted_var <- pmax(predicted_var, 1e-9) 
  actual_var <- pmax(actual_var, 1e-9)
  return(mean(actual_var / predicted_var - log(actual_var / predicted_var) - 1, na.rm = TRUE))
}
```


```{r}
#The only part to edit

predict_horizon_t <- 300
cutoff_point <- 600 - predict_horizon_t


# Avaliable Options: "GARCH11", "MLR", "RandomForest"
# Only one model at a time
selected_model <- "MLR" 

# Avaliable Options: "ALL" or c("seconds_elapsed", "n_obs", "last_return", "abs_last_return", "vol_trend", "RV_5", "RV_10", "RV_20", "RV_30", "RV_45", "RV_60", "RV_90", "RV_120", "RV_180", "RV_240", "RV_300", "ret_autocorr_30", "ret_autocorr_60", "ret_autocorr_120", "ret_autocorr_300", "EWMA_07", "EWMA_08", "EWMA_09", "EWMA_094", "EWMA_097", "EWMA_0985", "mean_spread_30", "mean_imbalance_30", "mean_depth_30", "book_pressure_30", "RV_ratio_5_30", "RV_ratio_30_300", "RV_ratio_60_300")

user_features_selection <- "ALL"

# Avaliable Options: "RMSE", "QLIKE"
# Free for all
selected_metrics <- c("RMSE", "QLIKE")
```

```{r}
ds_3_18 <- ds %>% filter(stock_id %in% c(3, 18)) %>% collect()

full_grid <- expand.grid(
  stock_id = unique(ds_3_18$stock_id),
  time_id = unique(ds_3_18$time_id),
  seconds_in_bucket = 0:599
)

ds_ret <- full_grid %>%
  left_join(ds_3_18, by = c("stock_id", "time_id", "seconds_in_bucket")) %>%
  group_by(stock_id, time_id) %>%
  arrange(seconds_in_bucket, .by_group = TRUE) %>%
  # Jupyter's ffill() counterpart
  mutate(across(starts_with("bid") | starts_with("ask"), ~na.locf(.x, na.rm = FALSE))) %>%
  # Jupyter's weighted_price and log_price_diff logic
  mutate(
    wap = (bid_price1 * ask_size1 + ask_price1 * bid_size1 + bid_price2 * ask_size2 + ask_price2 * bid_size2) / 
          (bid_size1 + ask_size1 + bid_size2 + ask_size2),
    log_ret = log(wap / lag(wap)) * 10000
  ) %>%
  mutate(log_ret = ifelse(is.na(log_ret), 0, log_ret)) %>%
  ungroup()

target <- ds_ret %>%
  filter(seconds_in_bucket > cutoff_point) %>%
  group_by(stock_id, time_id) %>%
  summarise(vol_second_half = mean(log_ret^2, na.rm = TRUE), .groups = "drop")
```

```{r}
predictor <- ds_ret %>%
  filter(seconds_in_bucket <= cutoff_point) %>%
  group_by(stock_id, time_id) %>%
  summarise(
    extract_features(log_ret, bid_price1, ask_price1, bid_size1, ask_size1, bid_price2, ask_price2, bid_size2, ask_size2),
    .groups = "drop"
  )

# Join and filter
data_ml <- inner_join(predictor, target, by = c("stock_id", "time_id"))
all_generated_feats <- setdiff(names(data_ml), c("stock_id", "time_id", "vol_second_half"))
feature_cols <- if(identical(user_features_selection, "ALL")) all_generated_feats else user_features_selection

data_ml <- data_ml[, c("stock_id", "time_id", "vol_second_half", feature_cols)] %>% na.omit()

set.seed(3182004)
unique_times <- unique(data_ml$time_id)
folds <- sample(rep(1:5, length.out = length(unique_times)))
data_ml <- data_ml %>% left_join(data.frame(time_id = unique_times, fold = folds), by = "time_id")
```


```{r, eval = (selected_model == "GARCH11")}
spec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0), include.mean = FALSE),
  # Adding fixed starting params can help convergence in small samples
  fixed.pars = list(omega = 1e-6) 
)

garch_results <- ds_ret %>%
  filter(seconds_in_bucket <= cutoff_point) %>%
  group_by(stock_id, time_id) %>%
  group_map(~{
    # Switch solver to 'gosolnp' or 'solnp' for better stability if 'hybrid' fails
    fit <- try(ugarchfit(spec, .x$log_ret, solver = 'solnp', 
                         fit.control = list(scale = 1)), silent = TRUE)
    
    # Check if fit failed OR if it didn't converge
    if(inherits(fit, "try-error") || convergence(fit) != 0) {
      return(data.frame(stock_id = .y$stock_id, time_id = .y$time_id, pred = NA))
    }
    
    fore <- ugarchforecast(fit, n.ahead = predict_horizon_t)
    # Get the forecasted variance
    f_sig <- as.numeric(sigma(fore))^2
    data.frame(stock_id = .y$stock_id, time_id = .y$time_id, pred = mean(f_sig))
  }, .keep = TRUE) %>%
  bind_rows()

garch_data <- inner_join(target, garch_results, by = c("stock_id", "time_id")) %>% na.omit()
final_predictions <- garch_data$pred
final_actuals <- garch_data$vol_second_half
```


```{r, eval = (selected_model == "MLR")}
#MLR

feature_cols <- setdiff(names(data), c("stock_id", "time_id", "fold", "vol_second_half"))
formula_str <- paste("vol_second_half ~", paste(feature_cols, collapse = " + "))

final_predictions <- numeric(nrow(data))
final_actuals <- numeric(nrow(data))

for (i in 1:5) {
  train_data <- data %>% filter(fold != i)
  test_data  <- data %>% filter(fold == i)
  
  model_mlr <- lm(as.formula(formula_str), data = train_data)
  preds <- predict(model_mlr, newdata = test_data)
  preds <- pmax(preds, EPS)
  
  final_predictions[data$fold == i] <- preds
  final_actuals[data$fold == i] <- test_data$vol_second_half
}
```

```{r, eval = (selected_model == "RandomForest")}
#RF

feature_cols <- setdiff(names(data), c("stock_id", "time_id", "fold", "vol_second_half"))
formula_str <- paste("vol_second_half ~", paste(feature_cols, collapse = " + "))

final_predictions <- numeric(nrow(data))
final_actuals <- numeric(nrow(data))

for (i in 1:5) {
  train_data <- data %>% filter(fold != i)
  test_data  <- data %>% filter(fold == i)
  
  model_ran <- ranger(
    formula = as.formula(formula_str), 
    data = train_data, 
    num.trees = 500
  )
  
  predicted_ran <- as.vector(predict(model_ran, data = test_data)$predictions)
  predicted_ran <- pmax(predicted_ran, EPS)
  
  final_predictions[data$fold == i] <- predicted_ran
  final_actuals[data$fold == i] <- test_data$vol_second_half
}
```

```{r, eval = ("RMSE" %in% selected_metrics)}
score_rmse <- pm_rmse(final_actuals, final_predictions)
cat("=== RMSE PERFORMANCE ===\n")
cat("Model:", selected_model, "\n")
cat("RMSE Score:", score_rmse, "\n")
```

```{r, eval = ("QLIKE" %in% selected_metrics)}
score_qlike <- pm_qlike(final_actuals, final_predictions)
cat("=== QLIKE PERFORMANCE ===\n")
cat("Model:", selected_model, "\n")
cat("QLIKE Score:", score_qlike, "\n")
```
