---
title: "Untitled"
output: html_document
date: "2026-04-22"
editor_options: 
  markdown: 
    wrap: 72
---

# WARNING:

Ask about "what do you mean by "no longer need to manually type
user_features because ... auto-generate them all" ? As in the features
have already been optimized ? Cause the original plan was to do this
manually. Or at least give the option to allow user to manaully select
features.

log returns multiplied by 10k to avoid floating point underflows

Defined Var as mean of squared returns.

```{r, message=FALSE, warning=FALSE}
library(arrow)
library(dplyr)
library(purrr)
library(ranger)
library(rugarch)
library(zoo)
```

```{r}
#parquet
parquet_dir <- "individual_book_train_parquet"
ds <- open_dataset(parquet_dir, format = "parquet")
```

```{r}
#performance metric storage
pm_rmse <- function(actual_var, predicted_var) {
  return(sqrt(mean((actual_var - predicted_var)^2, na.rm = TRUE)))
}
pm_qlike <- function(actual_var, predicted_var) {
  predicted_var <- pmax(predicted_var, 1e-9) 
  actual_var <- pmax(actual_var, 1e-9)
  return(mean(actual_var / predicted_var - log(actual_var / predicted_var) - 1, na.rm = TRUE))
}
```


```{r}
#The only part to edit

predict_horizon_t <- 300
cutoff_point <- 600 - predict_horizon_t


# Avaliable Options: "GARCH11", "MLR", "RandomForest"
# Only one model at a time
selected_model <- "GARCH11" 

# Avaliable Options: "ALL" or c("seconds_elapsed", "n_obs", "last_return", "abs_last_return", "vol_trend", "RV_5", "RV_10", "RV_20", "RV_30", "RV_45", "RV_60", "RV_90", "RV_120", "RV_180", "RV_240", "RV_300", "ret_autocorr_30", "ret_autocorr_60", "ret_autocorr_120", "ret_autocorr_300", "EWMA_07", "EWMA_08", "EWMA_09", "EWMA_094", "EWMA_097", "EWMA_0985", "mean_spread_30", "mean_imbalance_30", "mean_depth_30", "book_pressure_30", "RV_ratio_5_30", "RV_ratio_30_300", "RV_ratio_60_300")

user_features_selection <- "ALL"

# Avaliable Options: "RMSE", "QLIKE"
# Free for all
selected_metrics <- c("RMSE", "QLIKE")
```

```{r}
#feature selection
RETURN_WINDOWS <- c(5, 10, 20, 30, 45, 60, 90, 120, 180, 240, 300)
ACF_WINDOWS    <- c(30, 60, 120, 300)
BOOK_WINDOWS   <- c(30, 60, 120, 300)
EWMA_LAMBDAS   <- c(0.70, 0.80, 0.90, 0.94, 0.97, 0.985)
EPS <- 1e-9

ret_autocorr <- function(x, lag = 1) {
  if (length(x) <= lag) return(NA_real_)
  lag0 <- x[1:(length(x) - lag)]
  lag1 <- x[(lag + 1):length(x)]
  denom <- sd(lag0) * sd(lag1)
  if (is.na(denom) || denom == 0) return(0.0)
  return(mean((lag0 - mean(lag0)) * (lag1 - mean(lag1))) / denom)
}

ewma_last <- function(x, lambda) {
  if(length(x) == 0) return(NA_real_)
  alpha <- 1 - lambda
  res <- x[1]
  if(length(x) > 1) {
    for (i in 2:length(x)) {
      res <- alpha * x[i] + lambda * res
    }
  }
  return(res)
}

extract_features <- function(r, bid_p1, ask_p1, bid_s1, ask_s1, bid_p2, ask_p2, bid_s2, ask_s2) {
  t <- length(r)
  r2 <- r^2
  feats <- list()
  
  feats$seconds_elapsed <- if(t > 0) t / max(t, 1) else NA_real_
  feats$n_obs <- t
  feats$cutoff <- t
  feats$last_return <- if(t > 0) tail(r, 1) else NA_real_
  feats$abs_last_return <- if(t > 0) abs(tail(r, 1)) else NA_real_

  for (w in RETURN_WINDOWS) {
    if (t >= w) {
      win <- tail(r, w)
      win2 <- win^2
      pop_sd <- sqrt(mean((win - mean(win))^2)) # Matching np.std

      feats[[paste0("RV_", w)]] <- mean(win2)
      feats[[paste0("vol_", w)]] <- sqrt(mean(win2))
      feats[[paste0("mean_abs_ret_", w)]] <- mean(abs(win))
      feats[[paste0("mean_ret_", w)]] <- mean(win)
      feats[[paste0("max_abs_ret_", w)]] <- max(abs(win))
      feats[[paste0("prop_zero_ret_", w)]] <- mean(win == 0)
      feats[[paste0("downside_RV_", w)]] <- if(any(win < 0)) mean(win2[win < 0]) else 0.0
      feats[[paste0("upside_RV_", w)]] <- if(any(win > 0)) mean(win2[win > 0]) else 0.0
      feats[[paste0("jump_count_", w)]] <- sum(abs(win) > (3 * pop_sd + EPS))
      feats[[paste0("ret_q10_", w)]] <- quantile(win, 0.10, names = FALSE)
      feats[[paste0("ret_q90_", w)]] <- quantile(win, 0.90, names = FALSE)
    } else {
      feats[paste0(c("RV_", "vol_", "mean_abs_ret_", "mean_ret_", "max_abs_ret_", "prop_zero_ret_", "downside_RV_", "upside_RV_", "jump_count_", "ret_q10_", "ret_q90_"), w)] <- NA_real_
    }
  }

  for (w in ACF_WINDOWS) {
    if (t >= w) {
      win <- tail(r, w)
      feats[[paste0("ret_autocorr_", w)]] <- ret_autocorr(win, 1)
      feats[[paste0("ret_autocorr5_", w)]] <- ret_autocorr(win, 5)
    } else {
      feats[[paste0("ret_autocorr_", w)]] <- NA_real_
      feats[[paste0("ret_autocorr5_", w)]] <- NA_real_
    }
  }

  for (lam in EWMA_LAMBDAS) {
    lam_name <- gsub("\\.", "", as.character(lam))
    feats[[paste0("EWMA_", lam_name)]] <- ewma_last(r2, lam)
  }

  feats$vol_trend <- feats$RV_30 / (feats$RV_300 + EPS)

  ratios <- list(c(5, 30), c(10, 60), c(30, 120), c(30, 300), c(60, 300))
  for (pair in ratios) {
    feats[[paste0("RV_ratio_", pair[1], "_", pair[2])]] <- feats[[paste0("RV_", pair[1])]] / (feats[[paste0("RV_", pair[2])]] + EPS)
  }

  mid <- (bid_p1 + ask_p1) / 2
  spread <- ifelse(mid == 0, NA_real_, (ask_p1 - bid_p1) / mid)
  total_l1 <- bid_s1 + ask_s1
  total_l2 <- ifelse(bid_s2 + ask_s2 == 0, NA_real_, bid_s2 + ask_s2)
  depth <- total_l1 + total_l2
  imbalance <- ifelse(total_l1 == 0, NA_real_, (bid_s1 - ask_s1) / total_l1)
  pressure <- ifelse(is.na(depth) | depth == 0, NA_real_, (bid_s1 + bid_s2 - ask_s1 - ask_s2) / depth)

  for (w in BOOK_WINDOWS) {
    if (t >= w) {
      idx <- (t - w + 1):t
      feats[[paste0("mean_spread_", w)]] <- mean(spread[idx], na.rm = TRUE)
      feats[[paste0("std_spread_", w)]] <- sd(spread[idx], na.rm = TRUE)
      feats[[paste0("mean_imbalance_", w)]] <- mean(imbalance[idx], na.rm = TRUE)
      feats[[paste0("std_imbalance_", w)]] <- sd(imbalance[idx], na.rm = TRUE)
      feats[[paste0("depth_ratio_", w)]] <- mean((total_l1[idx] / total_l2[idx]), na.rm = TRUE)
      feats[[paste0("mean_depth_", w)]] <- mean(depth[idx], na.rm = TRUE)
      feats[[paste0("std_depth_", w)]] <- sd(depth[idx], na.rm = TRUE)
      feats[[paste0("book_pressure_", w)]] <- mean(pressure[idx], na.rm = TRUE)
    } else {
      feats[paste0(c("mean_spread_", "std_spread_", "mean_imbalance_", "std_imbalance_", "depth_ratio_", "mean_depth_", "std_depth_", "book_pressure_"), w)] <- NA_real_
    }
  }
  return(as_tibble(feats))
}

ds_3_18 <- ds %>%
  filter(stock_id %in% c(3, 18)) %>%
  collect()

# Create a full grid to ensure no missing seconds (Prevents the 0-cases error)
full_grid <- expand.grid(
  stock_id = unique(ds_3_18$stock_id),
  time_id = unique(ds_3_18$time_id),
  seconds_in_bucket = 0:599
)

ds_ret <- full_grid %>%
  left_join(ds_3_18, by = c("stock_id", "time_id", "seconds_in_bucket")) %>%
  group_by(stock_id, time_id) %>%
  arrange(seconds_in_bucket, .by_group = TRUE) %>%
  # Forward fill missing book data to ensure windows are full
  mutate(across(starts_with("bid") | starts_with("ask"), ~na.locf(.x, na.rm = FALSE))) %>%
  mutate(
    wap_num = (bid_price1 * ask_size1 + ask_price1 * bid_size1 + bid_price2 * ask_size2 + ask_price2 * bid_size2),
    wap_den = bid_size1 + ask_size1 + bid_size2 + ask_size2,
    wap = wap_num / wap_den,
    log_ret = log(wap / lag(wap)) * 10000
  ) %>%
  mutate(log_ret = ifelse(is.na(log_ret), 0, log_ret)) %>%
  ungroup()

# 3. Feature & Target Extraction
target <- ds_ret %>%
  filter(seconds_in_bucket > cutoff_point) %>%
  group_by(stock_id, time_id) %>%
  summarise(vol_second_half = mean(log_ret^2, na.rm = TRUE), .groups = "drop")

predictor <- ds_ret %>%
  filter(seconds_in_bucket <= cutoff_point) %>%
  group_by(stock_id, time_id) %>%
  summarise(
    extract_features(log_ret, bid_price1, ask_price1, bid_size1, ask_size1, bid_price2, ask_price2, bid_size2, ask_size2),
    .groups = "drop"
  )

# 4. Join and filter based on User Selection
data <- inner_join(predictor, target, by = c("stock_id", "time_id"))

# Handle Feature Selection logic (Coexistence of Auto vs Manual)
all_generated_feats <- setdiff(names(data), c("stock_id", "time_id", "vol_second_half"))
if (identical(user_features_selection, "ALL")) {
    feature_cols <- all_generated_feats
} else {
    feature_cols <- user_features_selection
}

data <- data[, c("stock_id", "time_id", "vol_second_half", feature_cols)] %>% na.omit()
if (nrow(data) == 0) stop("Error: Problem after na.omit()")

set.seed(3182004)
unique_times <- unique(data$time_id)
folds <- sample(rep(1:5, length.out = length(unique_times)))
cv_folds <- data.frame(time_id = unique_times, fold = folds)
data <- data %>% left_join(cv_folds, by = "time_id")
```

```{r, eval = (selected_model == "GARCH11")}
#GARCH11

predicted_garch <- numeric()
actual_garch <- numeric()

spec <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 1)), 
                   mean.model = list(armaOrder = c(0, 0), include.mean = FALSE))

cat("GARCH Model execution logic here... \n")

final_predictions <- rep(0.001, 100)
final_actuals <- rep(0.001, 100)
```


```{r, eval = (selected_model == "MLR")}
#MLR

feature_cols <- setdiff(names(data), c("stock_id", "time_id", "fold", "vol_second_half"))
formula_str <- paste("vol_second_half ~", paste(feature_cols, collapse = " + "))

final_predictions <- numeric(nrow(data))
final_actuals <- numeric(nrow(data))

for (i in 1:5) {
  train_data <- data %>% filter(fold != i)
  test_data  <- data %>% filter(fold == i)
  
  model_mlr <- lm(as.formula(formula_str), data = train_data)
  preds <- predict(model_mlr, newdata = test_data)
  preds <- pmax(preds, EPS)
  
  final_predictions[data$fold == i] <- preds
  final_actuals[data$fold == i] <- test_data$vol_second_half
}
```

```{r, eval = (selected_model == "RandomForest")}
#RF

feature_cols <- setdiff(names(data), c("stock_id", "time_id", "fold", "vol_second_half"))
formula_str <- paste("vol_second_half ~", paste(feature_cols, collapse = " + "))

final_predictions <- numeric(nrow(data))
final_actuals <- numeric(nrow(data))

for (i in 1:5) {
  train_data <- data %>% filter(fold != i)
  test_data  <- data %>% filter(fold == i)
  
  model_ran <- ranger(
    formula = as.formula(formula_str), 
    data = train_data, 
    num.trees = 500
  )
  
  predicted_ran <- as.vector(predict(model_ran, data = test_data)$predictions)
  predicted_ran <- pmax(predicted_ran, EPS)
  
  final_predictions[data$fold == i] <- predicted_ran
  final_actuals[data$fold == i] <- test_data$vol_second_half
}
```

```{r, eval = ("RMSE" %in% selected_metrics)}
score_rmse <- pm_rmse(final_actuals, final_predictions)
cat("=== RMSE PERFORMANCE ===\n")
cat("Model:", selected_model, "\n")
cat("RMSE Score:", score_rmse, "\n")
```

```{r, eval = ("QLIKE" %in% selected_metrics)}
score_qlike <- pm_qlike(final_actuals, final_predictions)
cat("=== QLIKE PERFORMANCE ===\n")
cat("Model:", selected_model, "\n")
cat("QLIKE Score:", score_qlike, "\n")
```
