---
title: "stock_27 modelling data3888"
format: html
editor: visual
---

```{r}
library(dplyr)
library(data.table)
library(arrow)
library(zoo)
library(stats)
library(ggplot2)

# load parquet
df <- read_parquet("/Users/vaniakumar/Desktop/data3888/data3888_volatility_modelling/individual_book_train_parquet/stock_27.parquet")
```

```{r}
length(unique(df$time_id))

dup_check <- df %>%
  group_by(time_id, seconds_in_bucket) %>%
  filter(n() > 1)

if (nrow(dup_check) > 0) {
  print("Duplicates found!")
} else {
  print("No duplicates: Each (time_id, seconds_in_bucket) pair is unique.")
}
```

```{r}
make_equally_spaced <- function(df) {
  all_time_ids <- unique(df$time_id)
  df_list <- list()
  
  for (tid in all_time_ids) {
    sub <- df %>%
      filter(time_id == tid) %>%
      arrange(seconds_in_bucket)
    
    full_index <- data.frame(seconds_in_bucket = 0:599)
    
    sub_full <- full_index %>%
      left_join(sub, by = "seconds_in_bucket") %>%
      mutate(time_id = tid) %>%
      arrange(seconds_in_bucket)
    
    # forward fill only
    sub_full <- sub_full %>%
      tidyr::fill(everything(), .direction = "down")
    
    df_list[[as.character(tid)]] <- sub_full
  }
  
  bind_rows(df_list)
}

df <- make_equally_spaced(df)
```

```{r}
df <- df %>%
  mutate(
    weighted_price = (
      bid_price1 * ask_size1 + ask_price1 * bid_size1 +
      bid_price2 * ask_size2 + ask_price2 * bid_size2
    ) / (bid_size1 + ask_size1 + bid_size2 + ask_size2)
  ) %>%
  group_by(time_id) %>%
  arrange(seconds_in_bucket) %>%
  mutate(
    log_price = log(weighted_price),
    log_price_diff = log_price - lag(log_price)
  ) %>%
  ungroup()
```

vol_proxy = (change in log(P))\^2 \
so that we can do ACF + PACF of squared returns

```{r}
library(stats)

max_lag <- 20
all_pacf <- list()
all_n <- c()

for (tid in unique(df$time_id)) {
  
  chunk <- df %>%
    filter(time_id == tid) %>%
    arrange(seconds_in_bucket)
  
  vol_proxy <- na.omit(chunk$log_price_diff)^2
  
  if (length(vol_proxy) < (max_lag + 1) || sd(vol_proxy) == 0) next
  
  pacf_vals <- pacf(vol_proxy, lag.max = max_lag, plot = FALSE)$acf
  
  all_pacf[[length(all_pacf) + 1]] <- pacf_vals
  all_n <- c(all_n, length(vol_proxy))
}

average_pacf <- Reduce("+", all_pacf) / length(all_pacf)
mean_n <- as.integer(mean(all_n))

conf_int <- ifelse(mean_n > 0, 1.96 / sqrt(mean_n), NA)
```

```{r}
pacf_df <- data.frame(
  lag = 0:max_lag,
  value = c(1, average_pacf)  # include lag 0
)

ggplot(pacf_df, aes(x = lag, y = value)) +
  geom_segment(aes(xend = lag, yend = 0)) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = conf_int, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -conf_int, linetype = "dashed", color = "red") +
  ggtitle("Average PACF of Squared Log Returns") +
  theme_minimal()
```

volatility is predictable in a linear in squares sense - we know that a model like **GARCH** or **stochastic volatility** can forecast next period’s variance better than a constant-variance assumption

```{r}
all_acf <- list()
all_n <- c()

for (tid in unique(df$time_id)) {
  
  chunk <- df %>%
    filter(time_id == tid) %>%
    arrange(seconds_in_bucket)
  
  vol_proxy <- na.omit(chunk$log_price_diff)^2
  
  if (length(vol_proxy) < (max_lag + 1) || sd(vol_proxy) == 0) next
  
  acf_vals <- acf(vol_proxy, lag.max = max_lag, plot = FALSE)$acf
  
  all_acf[[length(all_acf) + 1]] <- acf_vals
  all_n <- c(all_n, length(vol_proxy))
}

average_acf <- Reduce("+", all_acf) / length(all_acf)
mean_n <- as.integer(mean(all_n))

conf_int <- ifelse(mean_n > 0, 1.96 / sqrt(mean_n), NA)
```

```{r}
acf_df <- data.frame(
  lag = 0:max_lag,
  value = average_acf
)

ggplot(acf_df, aes(x = lag, y = value)) +
  geom_segment(aes(xend = lag, yend = 0)) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = conf_int, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -conf_int, linetype = "dashed", color = "red") +
  ggtitle("Average ACF of Squared Log Returns") +
  theme_minimal()
```

The slowly decaying ACF rules out simple i.i.d. returns (no clustering).

overall:\

|                                                   |                                  |
|---------------------------------------------------|----------------------------------|
| Volatility clustering positive & slowly decaying? | Yes                              |
| Volatility predictable?                           | Yes                              |
| Modelling approach justified?                     | Yes (for GARCH/ARCH-type models) |

**Modelling**

Feature Engineering

```{r}
library(dplyr)

# --- BASIC PREP ---
df <- df %>%
  mutate(
    mid_price = (bid_price1 + ask_price1) / 2,
    spread = ask_price1 - bid_price1,
    imbalance = (bid_size1 - ask_size1) / (bid_size1 + ask_size1 + 1e-6)
  ) %>%
  group_by(time_id) %>%
  arrange(seconds_in_bucket) %>%
  mutate(
    log_return = log(mid_price / lag(mid_price))
  ) %>%
  ungroup()

# --- ONLY FIRST HALF ---
df_first <- df %>%
  filter(seconds_in_bucket < 300)

# --- FEATURE ENGINEERING ---
features <- df_first %>%
  group_by(time_id) %>%
  summarise(
    
    # =========================
    # 🔥 VOLATILITY FEATURES
    # =========================
    rv_300 = sum(log_return^2, na.rm = TRUE),
    vol_300 = sqrt(rv_300),
    
    rv_60  = sum(log_return[seconds_in_bucket < 60]^2, na.rm = TRUE),
    rv_120 = sum(log_return[seconds_in_bucket < 120]^2, na.rm = TRUE),
    
    rv_ratio_60_300  = rv_60 / (rv_300 + 1e-8),
    rv_ratio_120_300 = rv_120 / (rv_300 + 1e-8),
    
    # =========================
    # 🔥 ROLLING VOL FEATURES (NEW)
    # =========================
    rolling_vol_mean = mean(abs(log_return), na.rm = TRUE),
    rolling_vol_sd   = sd(abs(log_return), na.rm = TRUE),
    rolling_vol_max  = max(abs(log_return), na.rm = TRUE),
    
    # =========================
    # 🔥 RETURN FEATURES
    # =========================
    return_mean = mean(log_return, na.rm = TRUE),
    return_sd   = sd(log_return, na.rm = TRUE),
    return_skew = mean((log_return - return_mean)^3, na.rm = TRUE),
    
    max_return = max(abs(log_return), na.rm = TRUE),
    
    # =========================
    # 🔥 SPREAD FEATURES
    # =========================
    avg_spread = mean(spread, na.rm = TRUE),
    sd_spread  = sd(spread, na.rm = TRUE),
    max_spread = max(spread, na.rm = TRUE),
    
    # =========================
    # 🔥 IMBALANCE FEATURES
    # =========================
    avg_imb = mean(imbalance, na.rm = TRUE),
    sd_imb  = sd(imbalance, na.rm = TRUE),
    
    imb_autocorr = cor(
      imbalance,
      dplyr::lag(imbalance),
      use = "complete.obs"
    ),
    
    # =========================
    # 🔥 VOLUME / LIQUIDITY
    # =========================
    avg_bid_size = mean(bid_size1, na.rm = TRUE),
    avg_ask_size = mean(ask_size1, na.rm = TRUE),
    volume_imbalance = mean(bid_size1 - ask_size1, na.rm = TRUE),
    
    # =========================
    # 🔥 PRICE FEATURES
    # =========================
    price_mean = mean(mid_price, na.rm = TRUE),
    price_sd   = sd(mid_price, na.rm = TRUE),
    
    # =========================
    # 🔥 MICROSTRUCTURE SIGNALS
    # =========================
    price_impact = cor(
      log_return,
      imbalance,
      use = "complete.obs"
    ),
    
    spread_return_corr = cor(
      spread,
      abs(log_return),
      use = "complete.obs"
    ),
    
    # =========================
    # 🔥 TIME STRUCTURE
    # =========================
    early_vol = sum(log_return[seconds_in_bucket < 150]^2, na.rm = TRUE),
    late_vol  = sum(log_return[seconds_in_bucket >= 150]^2, na.rm = TRUE),
    
    vol_trend = late_vol / (early_vol + 1e-8)
    
  ) %>%
  ungroup()

# =========================
# 🎯 TARGET (SECOND HALF)
# =========================
df_second <- df %>%
  filter(seconds_in_bucket >= 300) %>%
  group_by(time_id) %>%
  summarise(
    rv_second = sum(log_return^2, na.rm = TRUE)
  )

# =========================
# 🔥 FINAL DATASET
# =========================
final_df <- features %>%
  left_join(df_second, by = "time_id") %>%
  arrange(time_id) %>%
  mutate(
    target = log(rv_second + 1e-8),
    
    # =========================
    # 🔥 LAGGED FEATURES (NEW)
    # =========================
    lag_log_vol = lag(target),              # MOST IMPORTANT
    lag_vol     = lag(exp(target))          # optional (variance scale)
  ) %>%
  filter(!is.na(target))

# =========================
# ⚠️ CLEANING
# =========================
final_df[is.na(final_df)] <- 0
```

```{r}
# remove time_id (not a feature)
data_model <- final_df %>%
  select(-time_id)

# train/test split
set.seed(42)
n <- nrow(data_model)
train_idx <- sample(1:n, size = 0.8 * n)

train <- data_model[train_idx, ]
test  <- data_model[-train_idx, ]
```

baseline model

```{r}
lm_base <- lm(
  target ~ vol_300 + avg_spread,
  data = train
)

pred_base <- predict(lm_base, newdata = test)
```

```{r}
rmse_base <- sqrt(mean((test$target - pred_base)^2))

actual_var <- exp(test$target)
pred_var   <- exp(pred_base)

qlike_base <- mean((actual_var / pred_var) - log(actual_var / pred_var) - 1)
```

```{r}
rmse_base
qlike_base
```

used LASSO to shrink non-important features to 0

```{r}
full_model <- lm(target ~ ., data = train)

library(glmnet)

X_train <- as.matrix(train %>% select(-target))
y_train <- train$target

cv_model <- cv.glmnet(X_train, y_train, alpha = 1)

coef(cv_model, s = "lambda.min")

importance(rf_model)
```

```{r}
# -------------------------------
# LASSO REGRESSION
# -------------------------------
# Why:
# - Performs automatic feature selection
# - Reduces overfitting
# - Handles high-dimensional feature sets

library(glmnet)

X_train <- as.matrix(train %>% select(-target))
X_test  <- as.matrix(test %>% select(-target))

y_train <- train$target
y_test  <- test$target

# Cross-validated lambda (better than fixed)
cv_lasso <- cv.glmnet(X_train, y_train, alpha = 1)

pred_lasso <- predict(cv_lasso, newx = X_test, s = "lambda.min")
pred_lasso <- as.numeric(pred_lasso)

# Evaluation
rmse_lasso <- sqrt(mean((y_test - pred_lasso)^2))

actual_var <- exp(y_test)
pred_var   <- exp(pred_lasso)

qlike_lasso <- mean((actual_var / pred_var) - log(actual_var / pred_var) - 1)
```

```{r}
rmse_lasso
qlike_lasso
```

```{r}

# -------------------------------
# LINEAR REGRESSION
# -------------------------------
# Why:
# - Simple benchmark
# - Interpretable coefficients
# - Tests linear relationship assumption

lm_model <- lm(target ~ ., data = train)

pred_lm <- predict(lm_model, newdata = test)

# Evaluation
rmse_lm <- sqrt(mean((test$target - pred_lm)^2))

actual_var <- exp(test$target)
pred_var   <- exp(pred_lm)

qlike_lm <- mean((actual_var / pred_var) - log(actual_var / pred_var) - 1)
```

```{r}
rmse_lm
qlike_lm
```

```{r}
# -------------------------------
# RIDGE REGRESSION
# -------------------------------
# Why:
# - Keeps all features but shrinks coefficients
# - Handles multicollinearity better than OLS
# - More stable than LASSO

cv_ridge <- cv.glmnet(X_train, y_train, alpha = 0)

pred_ridge <- predict(cv_ridge, newx = X_test, s = "lambda.min")
pred_ridge <- as.numeric(pred_ridge)

# Evaluation
rmse_ridge <- sqrt(mean((y_test - pred_ridge)^2))

pred_var <- exp(pred_ridge)

qlike_ridge <- mean((actual_var / pred_var) - log(actual_var / pred_var) - 1)
```

```{r}
rmse_ridge
qlike_ridge
```

```{r}
# -------------------------------
# RANDOM FOREST
# -------------------------------
# Why:
# - Captures nonlinear relationships
# - Handles interactions automatically
# - Robust to noise and outliers

library(randomForest)

rf_model <- randomForest(
  target ~ .,
  data = train,
  ntree = 200
)

pred_rf <- predict(rf_model, newdata = test)

# Evaluation
rmse_rf <- sqrt(mean((test$target - pred_rf)^2))

pred_var <- exp(pred_rf)

qlike_rf <- mean((actual_var / pred_var) - log(actual_var / pred_var) - 1)
```

```{r}
rmse_rf
qlike_rf
```

```{r}
# -------------------------------
# DECISION TREE
# -------------------------------
# Why:
# - Simple nonlinear model
# - Easy to interpret splits
# - Good comparison vs Random Forest

library(rpart)

tree_model <- rpart(
  target ~ .,
  data = train,
  method = "anova",
  control = rpart.control(maxdepth = 6)
)

pred_tree <- predict(tree_model, newdata = test)

# Evaluation
rmse_tree <- sqrt(mean((test$target - pred_tree)^2))

pred_var <- exp(pred_tree)

qlike_tree <- mean((actual_var / pred_var) - log(actual_var / pred_var) - 1)
```

```{r}
rmse_tree
qlike_tree
```

```{r}
# -------------------------------
# XGBOOST
# -------------------------------

library(xgboost)

# Matrix format
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test)

# Train model
xgb_model <- xgboost::xgb.train(
  params = list(
    objective = "reg:squarederror",
    max_depth = 4,
    eta = 0.05,
    subsample = 0.8,
    colsample_bytree = 0.8
  ),
  data = dtrain,
  nrounds = 200,
  verbose = 0
)

# Predictions (log volatility)
pred_xgb <- predict(xgb_model, newdata = dtest)

# -------------------------------
# EVALUATION
# -------------------------------

# RMSE (log scale)
rmse_xgb <- sqrt(mean((y_test - pred_xgb)^2))

# Convert back to variance scale
actual_var <- exp(y_test)
pred_var   <- exp(pred_xgb)

# QLIKE
qlike_xgb <- mean((actual_var / pred_var) - log(actual_var / pred_var) - 1)

cat("RMSE:", rmse_xgb, "\n")
cat("QLIKE:", qlike_xgb, "\n")
```

```{r}
results <- data.frame(
  Model = c("Linear", "LASSO", "Ridge", "Random Forest", "Decision Tree", "XGBoost"),
  RMSE  = c(rmse_lm, rmse_lasso, rmse_ridge, rmse_rf, rmse_tree, rmse_xgb),
  QLIKE = c(qlike_lm, qlike_lasso, qlike_ridge, qlike_rf, qlike_tree, qlike_xgb)
)

print(results)
```

using PCA

```{r}
library(dplyr)

# -------------------------------
# 1. SPLIT DATA
# -------------------------------

X_train <- train %>% select(-target)
X_test  <- test  %>% select(-target)

y_train <- train$target
y_test  <- test$target

# -------------------------------
# 2. STANDARDISE FEATURES
# -------------------------------

X_mean <- apply(X_train, 2, mean)
X_sd   <- apply(X_train, 2, sd)

# avoid divide-by-zero
X_sd[X_sd == 0] <- 1

X_train_scaled <- scale(X_train, center = X_mean, scale = X_sd)
X_test_scaled  <- scale(X_test,  center = X_mean, scale = X_sd)

# -------------------------------
# 3. PCA (TRAIN ONLY)
# -------------------------------

pca_model <- prcomp(X_train_scaled, center = FALSE, scale. = FALSE)

# cumulative variance
var_explained <- cumsum(pca_model$sdev^2 / sum(pca_model$sdev^2))

# keep 90%–95% (more stable than strict 95%)
k <- which(var_explained >= 0.90)[1]

cat("Number of PCs used:", k, "\n")

# -------------------------------
# 4. TRANSFORM DATA
# -------------------------------

X_train_pca <- as.data.frame(pca_model$x[, 1:k])
X_test_pca  <- as.data.frame(predict(pca_model, newdata = X_test_scaled)[, 1:k])

# -------------------------------
# 5. FIT MODEL
# -------------------------------

lm_pca <- lm(y_train ~ ., data = X_train_pca)

pred_pca <- predict(lm_pca, newdata = X_test_pca)

# -------------------------------
# 6. SAFETY FIXES (IMPORTANT)
# -------------------------------

# remove NA / Inf
pred_pca[!is.finite(pred_pca)] <- median(pred_pca, na.rm = TRUE)

# cap extreme predictions (prevents QLIKE explosion)
pred_pca <- pmax(pmin(pred_pca, 5), -5)

# -------------------------------
# 7. EVALUATION
# -------------------------------

# RMSE
rmse_pca <- sqrt(mean((y_test - pred_pca)^2))

# QLIKE (stabilised)
actual_vol <- exp(y_test)

pred_vol <- exp(pmax(pred_pca, -5))   # prevents underflow

qlike_pca <- mean((actual_vol / pred_vol) -
                  log(actual_vol / pred_vol) - 1)

cat("PCA RMSE:", rmse_pca, "\n")
cat("PCA QLIKE:", qlike_pca, "\n")
```

```{r}
plot(var_explained, type = "l",
     main = "Cumulative Variance Explained",
     xlab = "Number of Components",
     ylab = "Variance Explained")
abline(h = 0.95, col = "red", lty = 2)
```

```{r}
loadings <- pca_model$rotation[, 1:5]  # first 5 PCs
print(loadings)
```

> PCA shows that dimensionality reduction alone is insufficient for volatility forecasting.
